---
output:
  pdf_document: default
  html_document: default
---
# basic equations for survival analysis and cox-regresion

More details can be found in my notebook https://zjuwhw.github.io/notebook-surivialR/
  
- Survival function, the probability of surviving up to a point t: $S(t)=pr(T>t), 0<t<\infty$
- Hazard function is the instantaneous failure rate: $h(t) = \lim\limits_{\delta \to \infty}\frac{pr(t<T<t+\delta|T>t)}{\delta}$
- Cumulative distribution function (CDF): $F(t) = pr(T \le t) = 1-S(t), 0<t<\infty$
- Probability density function (PDF): 
  - $f(t) = \frac{d}{dt} F(t) = \frac{d}{dt}[1-S(t)]=-\frac{d}{dt}S(t)$
  - $f(t) = h(t)S(t)$
- cumulative hazard function, the area under the hazard function up to time t:
  - $H(t)=\int_0^th(u)du$
  - $H(t) = -log[S(t)]$
  
Cox-regression

$$h_1(t)=\psi h_0(t); \psi = e^{z\beta}$$

- $\beta$: the effect size for the covariate
- $h_0(t)$: baseline hazard function; "baseline" means z=0
- $z$ is the covariate
- $z\beta$ is called "linear predictor" (log-hazard score)
- $\psi = e^{z\beta}$ is hazard ratio

- baseline cumulative hazard function: $H_0(t)=\int_0^th_0(t)dt$
- baseline survival function: $S_0(t)=exp[-H_0(t)]$
- survival function for a particular individual with covariate value z: $S(t|z)=[S_0(t)]^{exp(z\hat{\beta})}$
- absolute risk at time t with covariate value z: $1-S(t|z)$

### R code

##### toy example data

```{r}
### simple example with a dummy variable
library(survival)
tt = c(6, 7, 10, 15, 19, 25)
delta = c(1, 0, 1, 1, 0, 1)  ### 0 means censored; 1 means observed
summary(survfit(Surv(tt,delta)~1))
```


#### binary variable: baseline hazard function will be defined on the variable is 0

```{r}
trt = c(0, 0, 1, 0, 1, 1)  ## group 0 or 1
#### cox regression
result.cox = coxph(Surv(tt, delta)~trt, x=T, y=T)
summary(result.cox) ## beta is -1.3261

### no mean correction for binary variable
result.cox$means

#### linear predictor: beta*z
result.cox$linear.predictor
result.cox$coefficients*c(0,1)
predict(result.cox, type="lp")
predict(result.cox, newdata=data.frame(trt=c(0,1)), type="lp")  ## use the predict function with type="lp"

#### risk in the predict() function: exp(beta*z)
predict(result.cox, newdata=data.frame(trt=c(0,1)), type="risk")
exp(result.cox$linear.predictor)

### hazard ratio
exp(result.cox$coefficients)  ### HR<1 means decrease the risk

### baseline cumulative hazard function
#Be sure to use the option “centered = F” to cause it to estimate the cumulative hazard at
#beta= 0. The default is to estimate it at the mean of the covariates. This will often not
#make sense, particularly for categorical covariates such as treatment indicator, sex, or race
basehaz(result.cox, centered=F)
basehaz(result.cox, centered=T)
### no difference between centered=T and F return results in basehaz

### use the function survfit
base_surv = survfit(result.cox)
summary(base_surv)
base_hazard <- data.frame(
  time = base_surv$time,
  cumhaz = base_surv$cumhaz,
  surv = base_surv$surv
)
base_hazard 

### absolute risk at a time point: 1-S(t|z)
library(pec)
1-pec::predictSurvProb(result.cox, newdata=data.frame(trt=c(0,1)), time=25)
1 - base_hazard$surv[base_hazard$time==25]^(predict(result.cox, newdata=data.frame(trt=c(0,1)), type="risk")) 
```

##### continuous variable: baseline hazard function will be defined on the variable is at mean value


```{r}
age = c(50,30,50,34,65,90)
result.cox2 = coxph(Surv(tt, delta)~age,x=T, y=T)
summary(result.cox2)   ## coef: -0.03785

### mean correction
result.cox2$means
mean(age)

### linear predictor
result.cox2$linear.predictor
result.cox2$coefficients*(age-result.cox2$means)
predict(result.cox2, type="lp") 
predict(result.cox2, newdata = data.frame(age-mean(age)), type="lp") 

#### risk in the predict() function: exp(beta*z)
predict(result.cox2, type="risk") 
exp(predict(result.cox2, type="lp"))
predict(result.cox2, newdata = data.frame(age-mean(age)), type="lp") 

### hazard ratio
exp(result.cox2$coefficients)

### baseline cumulative hazard function
basehaz(result.cox2, centered=F) # age at 0
basehaz(result.cox2, centered=T) # age at the mean
basehaz(result.cox2, centered=F)$hazard * exp(mean(age)*result.cox2$coefficients)

### use the function survfit: same withe basehaz() with centered=T, default
base_surv2 = survfit(result.cox2)
summary(base_surv2)

base_hazard2 <- data.frame(
  time = base_surv2$time,
  cumhaz = base_surv2$cumhaz,
  surv = base_surv2$surv
)
base_hazard2 

### absolute risk at a time point: 1-S(t|z)
library(pec)

1-pec::predictSurvProb(result.cox2, newdata=data.frame(age=mean(age)), time=6)
1 - base_hazard2$surv[base_hazard2$time==6]^(predict(result.cox2, newdata=data.frame(age=mean(age)), type="risk"))


1-pec::predictSurvProb(result.cox2, newdata=data.frame(age=50), time=6)
1 - base_hazard2$surv[base_hazard2$time==6]^(predict(result.cox2, newdata=data.frame(age=50), type="risk"))
```